A Makefile for Developing Containerized \LaTeX\ Technical Documents

### 2020-12-13 ms.tex
improve ms.tex

\section{Introduction}
Developing technical documents that present both computationally generated results (variables, tables and figures) and natural text, is a central aspect of computational research.
One programming paradigm that was proposed as a solution is \textit{literate programming}~\cite{knuth1984literate}, in which the author inserts snippets of code and the output of its execution, alongside natural text.
Some applications of \textit{literate programming} include Jupyter~\cite{kluyver2016jupyter} which allows publishing code, results and explanations in a executable format, PythonTeX~\cite{poore2015pythontex} which allows Python to be executed in \LaTeX\ and ActivePapers~\cite{hinsen2014activepapers} in which code can be executed using the Java Virtual Machine.

@@

We propose a Makefile for developing containerized \LaTeX\ technical documents in POSIX-oriented operating systems.
By using portable technologies that withstood the test of time, we ensure the stability and portability of the proposed Makefile.
We write this paper\footnote{\url{https://github.com/pbizopoulos/a-makefile-for-developing-containerized-latex-technical-documents}}\footnote{\url{https://arxiv.org/abs/2005.12660}} using a template consisting the Makefile\footnote{\url{https://github.com/pbizopoulos/a-makefile-for-developing-containerized-latex-technical-documents-template}} and choose Docker as the container technology and the Python programming language~\cite{van2007python} for generating the results.

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[%
			grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
			edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
			\node[draw](document){document/}
			child{node[draw, anchor=west](makefile){Makefile}}
			child{node[draw, anchor=west, fill=green!20](tmp){tmp/}}
			child{node[draw, anchor=west, fill=red!20](tex){ms.tex}}
			child{node[draw, anchor=west, fill=red!20](bib){ms.bib}}
			child{node[draw, anchor=west, fill=blue!20](py){main.py}}
			child{node[draw, anchor=west, fill=blue!20](dockerfile){Dockerfile}}
			child{node[draw, anchor=west, fill=blue!20](requirements){requirements.txt}};
			\begin{scope}[on background layer]
				\node[draw, rounded corners, dashed, fill=gray!20, inner sep=4pt, fit=(document) (requirements)]{};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(py) (requirements)](code){};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(tex) (bib)](document){};
			\end{scope}
			\node[draw, right=1.5cm of code, fill=green!20] (results) {tmp/results};
			\path[draw, ->] (code) -- (results);
			\node[draw, right=4cm of document, fill=green!20] (pdf) {tmp/ms.pdf};
			\path[draw, ->] (results) |- (pdf);
			\path[draw, ->] (document) -- (pdf);
	\end{tikzpicture}
	\caption{The files of a Docker container template that uses the Makefile is depicted in the light gray background and the arrows denote the containerized data flow towards generating the \textit{document}.
	Green indicates temporary cache/generated files, red \textit{\LaTeX\ code} files, blue \textit{results code} files and within the dark gray background, files that are developed by the \textit{author}.}
	\label{fig:templatefiles}
\end{figure}

The three tools that are used by the Makefile are the Make program\footnote{\url{https://www.gnu.org/software/make/}} which is a build automation tool, \LaTeX~\cite{lamport1994latex} which is a typesetting system for publishing high quality research documents, and the container technology (such as Docker) which allows portable execution of \textit{code} between different hardware, operating systems(OSs) and environments.

@@

tmp/$(results): $(results_code)
	build image $(results_image)
	run container $(results_image)

@@

The proposed Makefile can be used for developing \LaTeX\ \textit{documents} that consist of \textit{results} generated from \textit{results code} executed within a container with a consistent (across different hardware, OS and programming language), minimal user interface as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Template make options.}]
make             # Generate the draft (fast) version.
make VER=--full  # Generate the full (slow) version.
make clean       # Remove the tmp/ directory.
\end{lstlisting}

@@

Regarding the \textit{results code} for Python, we can achieve deterministic stochasticity by setting the random seeds to a specific value such as:
\begin{lstlisting}[language=python, style=lststyle, caption={Python reproducibility commands for popular libraries.}]
# for build-in random module
random.seed(0)
# for numpy
np.random.seed(0)
# for tensorflow
tf.random.set_seed(0)
# for pytorch
torch.backends.cudnn.benchmark = False
torch.backends.cudnn.deterministic = True
torch.cuda.manual_seed_all(0)
torch.manual_seed(0)
\end{lstlisting}

Additionally an \textit{author} can use the POSIX utility \textit{cmp} to compare \textit{documents} (generated in a different session, OS or underlying hardware) byte by byte, and diffoscope\footnote{\url{https://diffoscope.org/}} for identifying sources of non-determinism as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Test draft document version reproducibility. This can also be used as a test script when pushing or pull requesting to a remote repository.}]
# Generate the first draft document version
make
# Backup the first resulting pdf
cp tmp/ms.pdf tmp/ms-previous.pdf
# Update the modification timestamp of main.py
touch main.py
# Generate the second draft document version
make
# Compare draft document versions byte by byte
cmp tmp/ms.pdf tmp/ms-previous.pdf
# Compare draft document versions
# in a human readable form
diffoscope tmp/ms.pdf tmp/ms-previous.pdf
\end{lstlisting}

@@

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/MNIST-loss.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/FashionMNIST-loss.pdf}}
	\caption{Comparison of MobilenetV2 with ReLU, ReLU6 and SiLU train and validation losses for MNIST and FashionMNIST.}
	\label{fig:loss}
\end{figure}

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/CIFAR10-ReLU-kernels.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/CIFAR100-ReLU-kernels.pdf}}
	\caption{The first 16 kernels of the first convolutional layer of MobilenetV2 with ReLU for CIFAR10 and CIFAR100.}
	\label{fig:kernels}
\end{figure}

@@

\section{Discussion}
The proposed Makefile is based on plain text file editing and can be used with any POSIX-oriented operating system that supports containers and Make.
POSIX and Make have existed for decades and have passed the test of time, while the container technology was standardized with the Open Container Initiative\footnote{\url{https://opencontainers.org/}} and there exist alternatives that could be used as drop-in replacements for Docker.
Moreover this Makefile can be combined with any version control system, container registry provider and text editor thus preventing `app/vendor lock-in' situations.

>>>>

\section{Introduction}
Developing technical documents that present both computationally generated results (variables, tables and figures) and natural text, is a central aspect of computational research.
One programming paradigm that was proposed as a solution is \textit{literate programming}~\cite{knuth1984literate}, in which the author inserts snippets of code and the output of its execution, alongside natural text.
Applications of \textit{literate programming} include Jupyter~\cite{kluyver2016jupyter} which allows publishing code, results and explanations in a executable format, PythonTeX~\cite{poore2015pythontex} which allows Python to be executed in \LaTeX\ and ActivePapers~\cite{hinsen2014activepapers} in which code can be executed using the Java Virtual Machine.

@@

We propose a Makefile for developing containerized \LaTeX\ technical documents in POSIX-oriented operating systems.
By using portable technologies that withstood the test of time, we ensure the stability and portability of the proposed Makefile.
We write this paper\footnote{\url{https://github.com/pbizopoulos/a-makefile-for-developing-containerized-latex-technical-documents}}\footnote{\url{https://arxiv.org/abs/2005.12660}} using a template that consists of the Makefile\footnote{\url{https://github.com/pbizopoulos/a-makefile-for-developing-containerized-latex-technical-documents-template}} and choose Docker as the container technology and the Python programming language~\cite{van2007python} for generating the results.

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[%
			grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
			edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
			\node[draw](document){document/}
			child{node[draw, anchor=west](makefile){Makefile}}
			child{node[draw, anchor=west, fill=green!20](tmp){tmp/}}
			child{node[draw, anchor=west, fill=red!20](tex){ms.tex}}
			child{node[draw, anchor=west, fill=red!20](bib){ms.bib}}
			child{node[draw, anchor=west, fill=blue!20](py){main.py}}
			child{node[draw, anchor=west, fill=blue!20](containerfile){Containerfile}}
			child{node[draw, anchor=west, fill=blue!20](requirements){requirements.txt}};
			\begin{scope}[on background layer]
				\node[draw, rounded corners, dashed, fill=gray!20, inner sep=4pt, fit=(document) (requirements)]{};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(py) (requirements)](code){};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(tex) (bib)](document){};
			\end{scope}
			\node[draw, right=1.5cm of code, fill=green!20] (results) {tmp/results};
			\path[draw, ->] (code) -- (results);
			\node[draw, right=4cm of document, fill=green!20] (pdf) {tmp/ms.pdf};
			\path[draw, ->] (results) |- (pdf);
			\path[draw, ->] (document) -- (pdf);
	\end{tikzpicture}
	\caption{The files of a template that uses the Makefile is depicted in the light gray background and the arrows denote the containerized data flow towards generating the \textit{document}.
	Green indicates temporary cache or generated files, red \textit{\LaTeX\ code} files, blue \textit{results code} files and within the dark gray background, files that are developed by the \textit{author}.}
	\label{fig:templatefiles}
\end{figure}

The proposed Makefile can be used for developing \textit{documents} that consist of \textit{results} generated from \textit{results code} executed within a container.
The three tools that are used by the Makefile are the Make program\footnote{\url{https://www.gnu.org/software/make/}} which is a build automation tool, \LaTeX~\cite{lamport1994latex} which is a typesetting system for publishing high quality research documents, and the container technology (such as Docker) which allows portable execution of \textit{code} between different hardware, operating systems and environments.

@@

tmp/$(results): $(results_code)
	build image $(results_image)
	run container $(results_image) $(VER)

@@

The Makefile consists of the following minimal user interface:
\begin{lstlisting}[language=bash, style=lststyle, caption={Template make options.}]
make             # Generate the draft (fast) version.
make VER=--full  # Generate the full (slow) version.
make clean       # Remove the tmp/ directory.
\end{lstlisting}

@@

Regarding the \textit{results code} for Python, we can achieve deterministic stochasticity by setting the random seeds to a specific value such as:
\begin{lstlisting}[language=python, style=lststyle, caption={Python reproducibility commands for popular libraries.}]
# build-in random module
random.seed(0)
# numpy
np.random.seed(0)
# tensorflow
tf.random.set_seed(0)
# pytorch
torch.backends.cudnn.benchmark = False
torch.backends.cudnn.deterministic = True
torch.cuda.manual_seed_all(0)
torch.manual_seed(0)
\end{lstlisting}

Additionally an \textit{author} can use the POSIX utility \textit{cmp} to compare \textit{documents} (generated in a different session, operating system or underlying hardware) byte by byte, and diffoscope\footnote{\url{https://diffoscope.org/}} for identifying sources of non-determinism as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Test draft document version reproducibility. This can also be used as a test script when pushing or pull requesting to a remote repository.}]
# Generate the first draft document version.
make
# Backup the first resulting pdf.
cp tmp/ms.pdf tmp/ms-previous.pdf
# Update the modification timestamp of main.py.
touch main.py
# Generate the second draft document version.
make
# Compare draft document versions byte by byte.
cmp tmp/ms.pdf tmp/ms-previous.pdf
# Compare draft document versions
# in a human readable form.
diffoscope tmp/ms.pdf tmp/ms-previous.pdf
\end{lstlisting}

@@

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/MNIST-loss.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/FashionMNIST-loss.pdf}}
	\caption{Comparison of the train and validation loss of MobilenetV2 with ReLU, ReLU6 and SiLU for MNIST and FashionMNIST.}
	\label{fig:loss}
\end{figure}

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/CIFAR10-ReLU-kernels.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/CIFAR100-ReLU-kernels.pdf}}
	\caption{The first 16 kernels of the first convolutional layer of MobilenetV2 with ReLU for CIFAR10 (left) and CIFAR100 (right).}
	\label{fig:kernels}
\end{figure}

@@

\section{Discussion}
The proposed Makefile is based on plain text file editing and can be used with any POSIX-oriented operating system that supports containers and Make.
POSIX and Make have existed for decades and have passed the test of time, while the container technology was standardized with the Open Container Initiative\footnote{\url{https://opencontainers.org/}} and there exist alternatives\footnote{\url{https://podman.io/}} that could be used as drop-in replacements for Docker.
Moreover the Makefile can be combined with any version control system, container registry provider and text editor thus preventing `app/vendor lock-in' situations.
